A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2024-11-09 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
Fix some typos before copying parts to NSF proposal

As mentioned in the introduction, our goal is to construct a probability theory of general applicability to physical systems. Our approach is not to define an axiomatic theory of probability and apply to physical systems, but rather characterize basic properties of physical states and from those build a theory of probability that fits those definitions. As it is common in our larger project, we start from axioms that, in our opinion, represent requirements of the scientific practice. The advantage is that the realm of applicability and physical interpretation will be clear from the start.

We start from requiring a physical theory to allow statistical descriptions: at the very least, it will provide us with a notion of ensemble, and the space of ensembles that are allowed by the theory. An ensemble can be understood as the infinite collection of the outputs of a particular preparation procedure. The ensemble space of classical mechanics, for example, is the space of continuous probability distributions of phase space. In quantum mechanics, instead, it is the space of density operators. While the ensemble space for each theory will be in principle different, they will all share common features: they must allow measurable properties of the system, probability distributions for outcomes of said properties, characterize reversible and irreversible processes through an entropy function, and so on. We are looking for necessary minimal basic requirements upon which these common tools can be built.

At least two arguments can be made to show that any physical theory must allow statistical descriptions, provide an ensemble space. First, note that physical laws are never about specific instances but rather regularities. They are statements of the type ``whenever we prepare this, we can measure that.'' This means that, ultimately, they are about ensembles. Second, if a physical theory is to be testable experimentally, and repeatedly so, it must be in terms of ensembles because this is what we prepare in practice and this is what, in the end, we characterize experimentally. The approach, then, can most likely be given firm conceptual grounds, which we leave as open but solvable philosophical problem.

@@

\subsection{Experimental verifiability}
The first requirement is that ensemble must be connected to experimental verification. That is, we must have enough \textbf{verifiable statements} at our disposal to define ensembles and tell them apart. By verifiable statement we mean a statement for which an experimental test is available that terminates in finite time if and only if the statement is true. For example, the statement ``the mass of the photon is less than $10^{-13} \, eV$'' is verifiable, while ``the mass of the photon is exactly $0 \, eV$'' is not verifiable due to its infinite precision. Based on previous work, this mathematically requires an ensemble space to be a $T_0$ second countable topological space, where open sets represent verifiable statements and Borel sets represent statements associated with a test, but without guarantee of termination.

The topological structure, then, represent the most foundational structure for a physical theory. It tells us why functions are ``well-behaved,'' which simply means they have to be topologically continuous in the ``natural topology'' induced by this requirement. It tells why probability is assigned to Borel sets (i.e. all statements associated with tests) and why the Banach-Tarski paradox does not apply in physics (i.e. non-Borel sets are physically ill-defined as they are not connected to experimental verification). Similarly, it tells us that sets with cardinality greater than that of the continuum are physically not relevant, as they cannot be given a $T_0$ second countable topology.

In the case of ensembles, typical verifiable statements will be ``the average energy of the particle is between 3 and 4 $eV$,'' ``the probability of getting heads is between 49 and 51 percent,'' ``the probability distribution for the position is a Gaussian of average 0 $m$ and 1 $m$ standard deviation within 1\%.'' Note that these are not results of single shot measurements, and therefore verifiable statements are distinctly different from observables in quantum mechanics. In the same manner, these are not measurements that extract one bit of information, as there is no requirement of termination in the negative case.

@@

\subsection{Statistical mixtures}
Another basic requirement is the ability to prepare a \textbf{mixture} of two ensembles. If $\ens[a], \ens[b] \in \Ens$ are two ensembles, and $p \in [0,1]$ a weight, then $\ens = p \ens[a] + (1-p) \ens[b] \in \Ens$ is the ensemble that describes a process that selects the first ensemble over the second $p$ percent of the times. Mathematically, the ensemble space is endowed with a \textbf{convex structure}. Moreover, since the mixing operation must be consistent with experimental verifiability, it will be topologically continuous and the ensemble space will be a topological convex space.

@@

The entropy is ultimately responsible for all geometrical structure in physics. In classical mechanics, the geometry is essentially defined by phase space volumes and areas in each DOF (degree of freedom). The entropy of uniform distributions is given by the logarithm of the volume, meaning that given the volumes we are able to calculate the entropy and given the entropy we are able to reconstruct the volume. Similarly, uniform marginals will recover DOF areas. In quantum mechanics, the inner product (i.e. the Born rule) allows us to calculate the entropy of the mixture of two pure states, and given the entropy we can recover the Born rule.

@@

On the convex structure we can define \textbf{pure states} as the ones that cannot be decomposed into other ensembles. For example, the pure states for a six-sided die are the ensemble perfectly prepared respectively on the six possible outcomes. All ensembles are convex combinations $\sum_{i=1}^{6} p_i \ens_i = 1$ of those pure states $\ens_i$. Mathematically, the pure states are the extreme points of the space and all other points can be understood as convex combinations over them. This is exactly how classical discrete spaces work and there is a temptation to think this is how everything works. For example, in classical mechanics the pure states would be perfect preparations of position and momentum, and ensembles are probability distributions. Not only this is problematic mathematically, it is conceptually not sustainable.

On physical grounds, we cannot consistently prepare classical particles perfectly, therefore there is no ensemble corresponding to a perfect preparation of a point of phase space. The mathematical facts that delta Dirac are not proper functions, their support has an empty interior, have minus infinite entropy, and so on, are simply a consequence of our conceptual problem. One may concede that the ensembles to do not exist physically, but may still want them to consider them as the limit of infinitesimal decomposition: classical ensembles can be decomposed into more and more precise ones, and the points of are the limit of this decomposition. Probability distributions are defined on these limits. This kind of works in classical mechanics, but in general.

>>>>

As mentioned in the introduction, our goal is to construct a probability theory of general applicability to physical systems. Our approach is not to define an axiomatic theory of probability and apply it to physical systems, but rather to characterize basic properties of physical states and from those build a theory of probability that fits those definitions. As is common in our larger project, we start from axioms that, in our opinion, represent requirements of scientific practice. The advantage is that the realm of applicability and physical interpretation will be clear from the start.

We start from requiring a physical theory to allow statistical descriptions: at the very least, it will provide us with a notion of ensemble, and the space of ensembles that are allowed by the theory. An ensemble can be understood as the infinite collection of the outputs of a particular preparation procedure. The ensemble space of classical mechanics, for example, is the space of continuous probability distributions on phase space. In quantum mechanics, instead, it is the space of density operators. While the ensemble space for each theory will in principle be different, they will all share common features: they must allow measurable properties of the system, probability distributions for outcomes of said properties, characterize reversible and irreversible processes through an entropy function, and so on. We are looking for necessary minimal basic requirements upon which these common tools can be built.

At least two arguments can be made to show that any physical theory must allow statistical descriptions, i.e.~provide an ensemble space. First, note that physical laws are never about specific instances but rather regularities. They are statements of the type ``whenever we prepare this, we can measure that.'' This means that, ultimately, they are about ensembles. Second, if a physical theory is to be testable experimentally, and repeatedly so, it must be in terms of ensembles because this is what we prepare in practice and this is what, in the end, we characterize experimentally. The approach, then, can most likely be given firm conceptual grounds, which we leave as an open but solvable philosophical problem.

@@

\subsection{Experimental verifiability}
The first requirement is that ensembles must be connected to experimental verification. That is, we must have enough \textbf{verifiable statements} at our disposal to define ensembles and tell them apart. By verifiable statement we mean a statement for which an experimental test is available that terminates in finite time if and only if the statement is true. For example, the statement ``the mass of the photon is less than $10^{-13} \, eV$'' is verifiable, while ``the mass of the photon is exactly $0 \, eV$'' is not verifiable due to its infinite precision. Based on previous work, this mathematically requires an ensemble space to be a $T_0$ second-countable topological space, where open sets represent verifiable statements and Borel sets represent statements associated with a test, but without guarantee of termination.

The topological structure, then, represents the most foundational structure for a physical theory. It tells us why functions are ``well-behaved,'' which simply means they have to be topologically continuous in the ``natural topology'' induced by this requirement. It tells why probability is assigned to Borel sets (i.e. all statements associated with tests) and why the Banach-Tarski paradox does not apply in physics (i.e. non-Borel sets are physically ill-defined as they are not connected to experimental verification). Similarly, it tells us that sets with cardinality greater than that of the continuum are not physically relevant, as they cannot be given a $T_0$ second-countable topology.

In the case of ensembles, typical verifiable statements will be ``the average energy of the particle is between 3 and 4 $eV$,'' ``the probability of getting heads is between 49 and 51 percent,'' ``the probability distribution for the position is a Gaussian of mean 0 $m$ and 1 $m$ standard deviation within 1\%.'' Note that these are not results of single-shot measurements, and therefore verifiable statements are distinctly different from observables in quantum mechanics. In the same manner, these are not measurements that extract one bit of information, as there is no requirement of termination in the negative case.

@@

\subsection{Statistical mixtures}
Another basic requirement is the ability to prepare a \textbf{mixture} of two ensembles. If $\ens[a], \ens[b] \in \Ens$ are two ensembles, and $p \in [0,1]$ a weight, then $\ens = p \ens[a] + (1-p) \ens[b] \in \Ens$ is the ensemble that describes a process that selects the first ensemble over the second $p$ percent of the times. Mathematically, the ensemble space is endowed with a \textbf{convex structure}. Moreover, since the mixing operation must be consistent with experimental verifiability, it will be topologically continuous, and the ensemble space will be a topological convex space.

@@

The entropy is ultimately responsible for all geometrical structure in physics. In classical mechanics, the geometry is essentially defined by phase space volumes and areas in each degree of freedom (DOF). The entropy of uniform distributions is given by the logarithm of the volume, meaning that given the volume we are able to calculate the entropy and given the entropy we are able to reconstruct the volume. Similarly, uniform marginals will recover DOF areas. In quantum mechanics, the inner product (i.e. the Born rule) allows us to calculate the entropy of the mixture of two pure states, and given the entropy we can recover the Born rule.

@@

On the convex structure we can define \textbf{pure states} as the ones that cannot be decomposed into other ensembles. For example, the pure states for a six-sided die are the ensembles perfectly prepared respectively on the six possible outcomes. All ensembles are convex combinations $\sum_{i=1}^{6} p_i \ens_i = 1$ of those pure states $\ens_i$. Mathematically, the pure states are the extreme points of the space and all other points can be understood as convex combinations over them. This is exactly how classical discrete spaces work and there is a temptation to think this is how everything works. For example, in classical mechanics the pure states would be perfect preparations of position and momentum, and ensembles are probability distributions. Not only is this problematic mathematically, it is conceptually not sustainable.

On physical grounds, we cannot consistently prepare classical particles perfectly, therefore there is no ensemble corresponding to a perfect preparation of a point of phase space. The mathematical facts that delta Dirac are not proper functions, their support has an empty interior, have minus infinite entropy, and so on, are simply a consequence of our conceptual problem. One may concede that the ensembles do not exist physically, but may still want them to consider them as the limit of infinitesimal decomposition: classical ensembles can be decomposed into more and more precise ones, and the points are the limit of this decomposition. Probability distributions are defined on these limits. This kind of works in classical mechanics, but in general.
